Supporting Information Appendix:\\ Black-swan events in animal populations

### 2017-01-07 text/supp.tex
Supplement tweeks

% Linux Libertine:
% \usepackage{textcomp}
% \usepackage[sb]{libertine}
% \usepackage[varqu,varl]{inconsolata}% sans serif typewriter
% \usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
% \usepackage[cal=boondoxo]{mathalfa} % mathcal
% %\useosf % osf for text, not math
% \usepackage[supstfm=libertinesups,%
%   supscaled=1.2,%
%   raised=-.13em]{superiors}

@@

\usepackage[round,sectionbib]{natbib}
\bibpunct{(}{)}{;}{a}{}{;}

@@

\title{Supporting Information:\\Black-swan events in animal
  populations}

@@

\item We removed all populations rated as $< 2$ in the GPDD quality assessment
  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
  \citep{sibly2005, ziebarth2010}.

\item Populations with negative abundance values were removed. Of the
  populations that remained at the end of our other filtering rules, the
  remaining populations with negative abundances listed were all from time
  series that had been standardized by subtracting the mean and dividing by the
  standard deviation. We verified this by locating the original papers the
  datasets were extracted from: \citet{colebrook1978} for zooplankton and
  \citet{lindstrom1995} for grouse. Since the papers did not include the
  original mean time-series values we could not back transform these data
  points.

@@

\item We filled in single recorded values of zero with the lowest non-zero
  value in the time series \citep{brook2006a}. This assumes that
  single values of zero result from abundance being low enough that censusing
  overlooked individuals that were actually present. We turned multiple zero
  values in a row into \texttt{NA} values. This implies that multiple zero
  values were either censusing errors or caused by emigration. Regardless, our
  population models were fit on a multiplicative (log) scale and so could not
  account for zero abundance. To avoid distorting the original data too
  strongly, we removed populations in which we filled in more than four zeros.

\item We removed all populations without at least four unique values
  \citep{brook2006a}.

@@

\item We then wrote an algorithm to find the longest unbroken window of
  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
  time series. If there were any populations with multiple windows of identical
  length, we took the most recent window. This is a longer window than used in
  some previous analyses \citep{brook2006a}, but since our model
  attempts to capture the shape of the distribution tails, our model requires
  more data.

@@

For the Gompertz model, our weakly informative priors (Fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by \citet{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
\citet{gelman2006c} and \citet{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \citep{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}C). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are concerned about when categorizing
populations as heavy tailed. We considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in \citet{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.
Furthermore, we considered the prior suggested in \citet{juarez2010} of
$\nu \sim \mathrm{Gamma}(2, 0.1)$, again limited to $\nu > 2$.
Compared to our main truncated exponential prior,
this prior places a higher probability density at $\nu < 10$ ($0.25$ vs.\ $0.08$)
and has a mode at $\nu = 10$.
This prior raised the estimated value of $\nu$
for some of the populations identified as heavy-tailed
with our default prior,
but overall gave us similar inference
with a Spearman correlation of 0.90 between Pr($\nu < 10$)
for the two priors.

@@

\subsection{Autocorrelated residuals} We considered a version of the Gompertz
model in which an autoregressive parameter was fit to the process-noise
residuals:
\begin{align}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model
estimates an additional parameter \(\phi\), which represents the correlation of
subsequent residuals. Based on the results of previous analyses with the
GPDD \citep{connors2014} and the chosen prior in previous
analyses \citep{thorson2014a}, we placed a moderately informative prior on
\(\phi\) that assumed the greatest probability density near zero with the
reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
\mathrm{max.} = 1)\).

\subsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
  \label{eq:rw-drift}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \citep{segura2013}.

@@

\subsection{Observation error} Observation error can bias parameter
estimates \citep{knape2012} and is known to affect the ability to detect extreme
events \citep{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \citep{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \citep{valpine2002, thorson2014b}.

@@

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
zero and twice the maximum observed abundance (similar and less informative
than in \citet{delean2013}). We set the prior on
\(r_\mathrm{max}\) as uniform between 0 and 20 as in \citet{delean2013}.
We used the same priors on \(\nu\) and \(\sigma\) as in
the Gompertz model.

@@

The probability density for the skewed Student-$t$ distribution
\citep{fernandez1998} equates to the following probability density as
implemented in \citet{king2012}:

@@

Since Stan only needs the probability density up to an additive
constant \citep{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student-$t$ distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

\section{Code availability.} Data and code to reproduce our analysis are
available at \url{https://github.com/seananderson/heavy-tails}.

@@

\bibliographystyle{apalike}\bibliography{jshort,newbib}

@@

%\begin{helvetica}
\singlespacing
\begin{scriptsize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{5.0cm}>{\RaggedRight}p{0.65cm}>{\RaggedRight}p{2.1cm}>{\RaggedRight}p{4.2cm}>{\RaggedRight}p{1.6cm}}
\caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model. Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted $t$ distribution.}\\

@@

\noindent
Example Stan code for a heavy-tailed Gompertz model. The specific code for
the various models in our analysis is available at
\url{https://github.com/seananderson/heavy-tails}.

%\begin{spacing}{1.15}
%% \begin{footnotesize}
%% \begin{verbatim}
%% data {
%%   int<lower=3> N;              // number of observations
%%   vector[N] y;                 // vector to hold ln abundance observations
%%   real<lower=0> nu_rate;       // rate parameter for nu exponential prior
%% }
%% parameters {
%%   real lambda;                 // Gompertz growth rate parameter
%%   real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
%%   real<lower=0> sigma_proc;    // process noise scale parameter
%%   real<lower=2> nu;            // t-distribution degrees of freedom
%%   real<lower=-1, upper=1> phi; // AR1 parameter
%%   vector[N] U;                 // unobserved states
%%   real<lower=0> sigma_obs;     // specified observation error SD
%% }
%% transformed parameters {
%%   vector[N] epsilon;           // error terms
%%   epsilon[1] <- 0;
%%   for (i in 2:N) {
%%     epsilon[i] <- U[i] - (lambda + b * U[i - 1])
%%                        - (phi * epsilon[i - 1]);
%%   }
%% }
%% model {
%%   // priors:
%%   nu ~ exponential(nu_rate);
%%   lambda ~ normal(0, 10);
%%   sigma_proc ~ cauchy(0, 2.5);
%%   phi ~ normal(0, 1);
%%   // data model:
%%   for (i in 2:N) {
%%     U[i] ~ student_t(nu,
%%                      lambda + b * U[i - 1]
%%                      + phi * epsilon[i - 1],
%%                      sigma_proc);
%%   }
%%   y ~ normal(U, sigma_obs);
%% }
%% \end{verbatim}
%% \end{footnotesize}
%%

\begin{footnotesize}
\begin{verbatim}
data {
  int<lower=1> N;              // number of observations
  vector[N] y;                 // vector to hold log abundance observations
  real<lower=0> nu_rate;       // rate parameter for nu exponential prior
}
parameters {
  real lambda;                 // Gompertz growth rate parameter
  real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
  real<lower=0> sigma_proc;    // process noise scale parameter
  real<lower=2> nu;            // t-distribution degrees of freedom
}
model {
  nu ~ exponential(nu_rate);
  lambda ~ normal(0, 10);
  sigma_proc ~ cauchy(0, 2.5);
  for (i in 2:N) {
    y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
  }
}
\end{verbatim}
\end{footnotesize}

\clearpage
\noindent
Stan code for the multilevel beta regression:
\begin{footnotesize}
\verbatiminput{../analysis/betareg4.stan}
\end{footnotesize}

>>>>

\widowpenalty10000
\clubpenalty10000

% Linux Libertine:
\usepackage{textcomp}
\usepackage[sb]{libertine}
\usepackage[varqu,varl]{inconsolata}% sans serif typewriter
\usepackage[libertine,bigdelims,vvarbb]{newtxmath} % bb from STIX
\usepackage[cal=boondoxo]{mathalfa} % mathcal
%\useosf % osf for text, not math
\usepackage[supstfm=libertinesups,%
  supscaled=1.2,%
  raised=-.13em]{superiors}

@@

% \usepackage[round,sectionbib]{natbib}
% \bibpunct{(}{)}{;}{a}{}{;}

@@

\title{Supporting Information Appendix:\\
  Black-swan events in animal populations}

@@

\item We removed all populations rated as $< 2$ in the GPDD quality assessment
  (on a scale of $1$ to $5$, with $1$ being the lowest quality data)
  \cite{sibly2005, ziebarth2010}.

\item Populations with negative abundance values were removed. Of the
  populations that remained at the end of our other filtering rules, the
  remaining populations with negative abundances listed were all from time
  series that had been standardized by subtracting the mean and dividing by the
  standard deviation. We verified this by locating the original papers the
  datasets were extracted from: Ref.~\cite{colebrook1978} for zooplankton and
  Ref.~\cite{lindstrom1995} for grouse. Since the papers did not include the
  original mean time-series values we could not back transform these data
  points.

@@

\item We filled in single recorded values of zero with the lowest non-zero
  value in the time series \cite{brook2006a}. This assumes that
  single values of zero result from abundance being low enough that censusing
  overlooked individuals that were actually present. We turned multiple zero
  values in a row into \texttt{NA} values. This implies that multiple zero
  values were either censusing errors or caused by emigration. Regardless, our
  population models were fit on a multiplicative (log) scale and so could not
  account for zero abundance. To avoid distorting the original data too
  strongly, we removed populations in which we filled in more than four zeros.

\item We removed all populations without at least four unique values
  \cite{brook2006a}.

@@

\item We then wrote an algorithm to find the longest unbroken window of
  abundance (no \texttt{NA}s) with at least $20$ time steps in each population
  time series. If there were any populations with multiple windows of identical
  length, we took the most recent window. This is a longer window than used in
  some previous analyses \cite{brook2006a}, but since our model
  attempts to capture the shape of the distribution tails, our model requires
  more data.

@@

For the Gompertz model, our weakly informative priors (Fig.~\ref{fig:priors}) were:
\begin{align}
b &\sim \mathrm{Uniform}(-1, 2)\\
\lambda &\sim \mathrm{Normal}(0, 10^2)\\
\sigma &\sim \mathrm{Half\mhyphen Cauchy} (0, 2.5)\\
\nu &\sim \mathrm{Truncated\mhyphen Exponential}(0.01, \mathrm{min.} = 2).
\end{align}
Our prior on \(b\) was uninformative between values of \(-1\) and \(2\). We
would not expect values of \(b\) with levels of density dependence as low as
\(-1\) (very strong inverse density dependence), nor would we generally expect
values above \(1\). We allowed values of \(b\) above \(1\) to allow for
non-stationary time series of growth rates. The estimates of \(b\) were well
within these bounds. Our prior on \(\lambda\) was weakly informative within the
range of expected values for population growth and is similar to the default
priors suggested by Ref.~\cite{gelman2008d} for intercepts of
regression models. Our half-Cauchy prior on \(\sigma\) follows
Ref.~\cite{gelman2006c} and Ref.~\cite{gelman2008d} and the specific scale parameter of
\(2.5\) is based on our expected range of the value in nature from previous
studies \cite{connors2014}. In our testing of a subsample of populations, our
parameter estimates were not qualitatively changed by switching to a uniform
prior on \(\sigma\), but the weakly informative half-Cauchy prior sped up chain
convergence.

To test if the prior on \(\nu\) influenced our estimate of black-swan dynamics,
we refit our models with weaker and stronger priors. Our base model used a
prior on \(\nu\) of Truncated-Exponential(0.01, min.\ = 2). For a weaker prior
we used Truncated-Exponential(0.005, min.\ = 2) and for a stronger prior we
used Truncated-Exponential(0.02, min.\ = 2) (Fig.~\ref{fig:priors}C). Note that
the base and weaker priors are relatively flat within the region of \(\nu <
20\), which is the region we are concerned about when categorizing
populations as heavy tailed. We considered using a Uniform(\(0, 1\))
prior on \(1/\nu\) as used in Ref.~\cite{gelman2014}, but this
prior puts considerably more weight on lower (heavy-tailed) values of \(\nu\)
than the truncated exponential priors considered here,
and the results are therefore not shown.
Furthermore, we considered the prior suggested in Ref.~\cite{juarez2010} of
$\nu \sim \mathrm{Gamma}(2, 0.1)$, again limited to $\nu > 2$.
Compared to our main truncated exponential prior,
this prior places a higher probability density at $\nu < 10$ ($0.25$ vs.\ $0.08$)
and has a mode at $\nu = 10$.
This prior raised the estimated value of $\nu$
for some of the populations identified as heavy-tailed
with our default prior,
but overall gave us similar inference
with a Spearman correlation of 0.90 between Pr($\nu < 10$)
for the two priors.

@@

\subsection{Autocorrelated residuals} We considered a version of the Gompertz
model in which an autoregressive parameter was fit to the process-noise
residuals:
\begin{align}
x_t &= \lambda + b x_{t-1} + \epsilon_t\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, \phi \epsilon_{t-1}, \sigma).
\end{align}
In addition to the parameters in the original Gompertz model, this model
estimates an additional parameter \(\phi\), which represents the correlation of
subsequent residuals. Based on the results of previous analyses with the
GPDD \cite{connors2014} and the chosen prior in previous
analyses \cite{thorson2014a}, we placed a moderately informative prior on
\(\phi\) that assumed the greatest probability density near zero with the
reduced possibility of \(\phi\) being near \(-1\) or \(1\). Specifically, we
chose \(\phi \sim \mathrm{truncated\mhyphen Normal}(0, 0.5, \mathrm{min.} = -1,
\mathrm{max.} = 1)\).

\subsection{Density independence} We fit a simplified version of the
Gompertz model in which the density dependence parameter \(b\) was fixed at
\(1\) (density independent). This is equivalent to fitting a random walk model
(with drift) to the \(\log\) abundances or assuming the growth rates are drawn
from a stationary distribution. The model was as follows:
\begin{align}
  \label{eq:rw-drift}
x_t &= \lambda + x_{t-1} + \epsilon_t\\
\epsilon &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma).
\end{align}
We fit this model for three reasons: (1) it is computationally simpler and so
provides a check that our more complicated full Gompertz model was obtaining
reasonable estimates of \(\nu\), (2) it provides a test of whether density
dependence was affecting our perception of heavy tails, (3) it
matches how some previous authors have modelled heavy tails without accounting
for density dependence \cite{segura2013}.

@@

\subsection{Observation error} Observation error can bias parameter
estimates \cite{knape2012} and is known to affect the ability to detect extreme
events \cite{ward2007}. In our main analysis, we fit a model that ignored
observation error. One way to account for observation error would be to fit a
full state-space model that simultaneously estimates the magnitude of process
noise and observation error. However, simultaneously estimating observation and
process noise is a challenging problem (e.g.\ because the observation and
process noise parameters tend to negatively covary in model fitting) and is
known to sometimes result in identifiability issues with the Gompertz
population model \cite{knape2008}. Furthermore, our model was applied to
hundreds of time series, often of short length (as few as \minTimeSteps\ time
steps) and our model estimates an additional parameter---the shape of the
process deviation tails---potentially making identifiability and computational
issues even greater. Therefore, we considered a version of the base Gompertz
model that allowed for a fixed level of observation error:
\begin{align}
U_t &= \lambda + b U_{t-1} + \epsilon_t\\
x_t &\sim \mathrm{Normal}(U_t, \sigma_\mathrm{obs}^2)\\
\epsilon_t &\sim \mathrm{Student}\mhyphen t(\nu, 0, \sigma_\mathrm{proc}),
\end{align}
where \(U\) represents the served state vector, \(\sigma_\mathrm{obs}\)
represents the standard deviation of observation error (on a log scale), and
\(\sigma_\mathrm{proc}\) represent the process noise scale parameter. We set
\(\sigma_\mathrm{obs}\) to \(0.2\), which represents a typical value used in
simulation analyses \cite{valpine2002, thorson2014b}.

@@

To fit the Ricker-logistic models, we chose a prior on \(K\) uniform between
zero and twice the maximum observed abundance (similar and less informative
than in Ref.~\cite{delean2013}). We set the prior on
\(r_\mathrm{max}\) as uniform between 0 and 20 as in Ref.~\cite{delean2013}.
We used the same priors on \(\nu\) and \(\sigma\) as in
the Gompertz model.

@@

The probability density for the skewed Student-$t$ distribution
\cite{fernandez1998} equates to the following probability density as
implemented in Ref.~\cite{king2012}:

@@

Since Stan only needs the probability density up to an additive
constant \cite{stan-manual2014}, we
can replace $\log(2 \gamma)$ in the above for $\log(\gamma)$. The final log
probability density for the skewed Student-$t$ distribution implemented in
Stan\footnote{Credit to Stan developer Bob Carpenter:
\url{https://groups.google.com/d/msg/stan-users/jeZQnQRUsDs/BeqhicWm1GEJ}}
looks like the following:

@@

\section{Code availability} Data and code to reproduce our analysis are
available at\\ \url{https://github.com/seananderson/heavy-tails}

@@

\bibliographystyle{pnas2011}\bibliography{jshort,newbib}

@@

%\begin{helvetica}
\singlespacing
\begin{scriptsize}
\begin{longtable}{>{\RaggedRight}m{1.4cm}>{\RaggedRight}p{5.3cm}>{\RaggedRight}p{0.65cm}>{\RaggedRight}p{1.5cm}>{\RaggedRight}p{4.2cm}>{\RaggedRight}p{1.6cm}}
\caption{Populations with a high probability of heavy-tailed dynamics in the base heavy-tailed Gompertz population dynamics model. Shown are the log abundance time series, population descriptions, Global Population Dynamics Database Main IDs, citation for the data source or separate verification literature, a description of the cause of the black-swan events (if known), the probability of heavy tails as calculated by our model, and median estimate of $\nu$ from our model with 90\% quantile credible intervals indicated in parentheses. Red dots on the time series indicate downward black-swan events and blue values indicate upward black-swan events that have a $10^{-4}$ probability or less of occurring if the population dynamics were explained by a Gompertz model with normally distributed process noise with a standard deviation equal to the scale parameter in the fitted $t$ distribution.}\\

@@

% \noindent
% Example Stan code for a heavy-tailed Gompertz model. The specific code for
% the various models in our analysis is available at
% \url{https://github.com/seananderson/heavy-tails}.

%\begin{spacing}{1.15}
%% \begin{footnotesize}
%% \begin{verbatim}
%% data {
%%   int<lower=3> N;              // number of observations
%%   vector[N] y;                 // vector to hold ln abundance observations
%%   real<lower=0> nu_rate;       // rate parameter for nu exponential prior
%% }
%% parameters {
%%   real lambda;                 // Gompertz growth rate parameter
%%   real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
%%   real<lower=0> sigma_proc;    // process noise scale parameter
%%   real<lower=2> nu;            // t-distribution degrees of freedom
%%   real<lower=-1, upper=1> phi; // AR1 parameter
%%   vector[N] U;                 // unobserved states
%%   real<lower=0> sigma_obs;     // specified observation error SD
%% }
%% transformed parameters {
%%   vector[N] epsilon;           // error terms
%%   epsilon[1] <- 0;
%%   for (i in 2:N) {
%%     epsilon[i] <- U[i] - (lambda + b * U[i - 1])
%%                        - (phi * epsilon[i - 1]);
%%   }
%% }
%% model {
%%   // priors:
%%   nu ~ exponential(nu_rate);
%%   lambda ~ normal(0, 10);
%%   sigma_proc ~ cauchy(0, 2.5);
%%   phi ~ normal(0, 1);
%%   // data model:
%%   for (i in 2:N) {
%%     U[i] ~ student_t(nu,
%%                      lambda + b * U[i - 1]
%%                      + phi * epsilon[i - 1],
%%                      sigma_proc);
%%   }
%%   y ~ normal(U, sigma_obs);
%% }
%% \end{verbatim}
%% \end{footnotesize}
%%
%%
%% \begin{footnotesize}
%% \begin{verbatim}
%% data {
%%   int<lower=1> N;              // number of observations
%%   vector[N] y;                 // vector to hold log abundance observations
%%   real<lower=0> nu_rate;       // rate parameter for nu exponential prior
%% }
%% parameters {
%%   real lambda;                 // Gompertz growth rate parameter
%%   real<lower=-1, upper=2> b;   // Gompertz density dependence parameter
%%   real<lower=0> sigma_proc;    // process noise scale parameter
%%   real<lower=2> nu;            // t-distribution degrees of freedom
%% }
%% model {
%%   nu ~ exponential(nu_rate);
%%   lambda ~ normal(0, 10);
%%   sigma_proc ~ cauchy(0, 2.5);
%%   for (i in 2:N) {
%%     y[i] ~ student_t(nu, lambda + b * y[i-1], sigma_proc);
%%   }
%% }
%% \end{verbatim}
%% \end{footnotesize}
%%

\clearpage
\noindent
Stan code for the multilevel beta regression:
\begin{scriptsize}
\verbatiminput{../analysis/betareg4.stan}
\end{scriptsize}
